---
title: "Programming: Functions, Loops, If/Else Statements"
output: 
  html_document:
    df_print: paged
    theme: cosmo
    highlight: haddock
    toc: yes
    toc_float: yes
    code_fold: show
    self-countained: false
---

<style>
body {
text-align: left}
</style>

<br>

## **Functions**  

An *AWESOME* feature of R, and perhaps a main reason for using it, is that it allows you to program your own functions. In fact, all the commands that you use in R *are* functions.  

We can look at the basic structure of a function in R by typing the function name, but excluding the parentheses. For example, lets take a look at the function for taking the row mean of an object, `rowMeans()`. Just type:
```{r,echo=TRUE,eval=FALSE}
rowMeans # There is a lot going on, but you see the structure.
```

The basic structure of a function is:  
`function.name <- function(`arguments or things you need separated by commas`){`  
  actions or stuff you want the function to be doing  
  `return(`information you want from the function`)`  
  `}`  

Basically, we are passing in an object or objects, doing something to it, and then returning an output.  

For example, let’s create a function that calculates the mean of a variable:  
```{r,echo=TRUE,eval=TRUE}
function.mean <- function( input ){
  our.mean <- sum( input )/length( input ) 
  return( our.mean ) 
}
```

This function takes an object `input` and divides the sum of input by the length of input and returns this ratio as an object called `our.mean`. Note that the names we use for the objects are arbitrary.  

The function reads as: create a function called `function.mean`, this function takes one argument called `input`, the function will create an object called `our.mean` by summing input and dividing it by the length of input, the function then returns the object `our.mean`.  

Remember that we can see our function by just typing the name `function.mean`.  

Now, let’s create an object called `x` and calculate the mean using our function:  
```{r,echo=TRUE,eval=TRUE}
x <- c( 1,2,3,4,5 )
function.mean( x )
```  

Note here that `x` is the input in the function.   

We can also create a object that is returned by the function.  
```{r,echo=TRUE,eval=TRUE}
x.mean <- function.mean( x )
x.mean
```
The command: `function.mean(x)` just returns the mean, but the command: `x.mean <- function.mean(x)` creates the object `x.mean` that is the mean of the object `x`.  

Commenting inside the function is helpful for keeping track of what is going on, as well as for others who might want to use a function you have created. In R, any line with a `#` sign in front is commented out of the code in the function and it will pass over it. For example:  
```{r,echo=TRUE,eval=TRUE}
function.mean <- function( input ){
  our.mean <- sum( input )/length( input ) #the equation for the mean.
  return( our.mean ) #the object 'our.mean' is what is created.
}
```
It is important to note that R will overwrite the names of existing functions. For example, the function `rowMeans()` can be overwritten by the command `rowMeans <- function(input){`...`}`. So be careful with your naming of functions. To be safe, just type the name of what you want to call your function to see if it exists in the loaded libraries.  
<br>

## **More about the function of the `function()` Function**  
The `return()` command within the `function()` function is helpful when our function creates multiple objects, but we only want particular objects returned by the function. For example, let’s create a function that takes the mean of two different inputs, then creates a new object that has both of these means as a single object:  
```{r,echo=TRUE,eval=TRUE}
function.two.means <- function( input1,input2 ){ #this function takes two arguments.
    mean.1 <- sum( input1 )/length( input1 ) # take the mean of the first object.
    mean.2 <- sum( input2 )/length( input2 ) # take the mean of the second object.
    both.means <- c( mean.1,mean.2 ) #combine the means into a object.
    return( both.means ) #return the object that is both means.
	}
```
Let’s give it a try:  
```{r,echo=TRUE,eval=TRUE}
x <- c( 1,2,3,4,5 )
y <- c( 6,7,8,9,10 )
```  

Create the mean of `x`:  
```{r,echo=TRUE,eval=TRUE}
function.mean( x )
```

Create the mean of `y`:  
```{r,echo=TRUE,eval=TRUE}
function.mean( y )
```  

Now, create the mean of the means. *Note* that we have to put in two arguments here:  
```{r,echo=TRUE,eval=TRUE}
function.two.means( x,y ) 
```

Note that this function creates an object with two means. That is to say, a *vector* with dimensions *1 x 2*. That is different then a function which takes the mean of the means. For example:  
```{r,echo=TRUE,eval=TRUE}
function.mean.of.means <- function( input1,input2 ){ #this function takes two arguments.
    mean.1 <- sum( input1 )/length( input1 ) # take the mean of the first object.
    mean.2 <- sum( input2 )/length( input2 ) # take the mean of the second object.
    mean.of.means <- mean( c( mean.1,mean.2 ) ) #take the mean of the means.
    return( mean.of.means ) #return the object that is the mean of both means.
	}
```
Let’s give it a try:  
```{r,echo=TRUE,eval=TRUE}
x <- c( 1,2,3,4,5 )
y <- c( 6,7,8,9,10 )
function.mean( x )
function.mean( y )
function.mean.of.means( x,y )
```

Now, compare that to:
```{r,echo=TRUE,eval=TRUE}
function.two.means( x,y )
```

*Question*: What is the difference between the statement `both.means <- c(mean.1,mean.2)` in the first function and the statement `mean.of.means <- mean(c(mean.1,mean.2))` in the second function?  
<br>

## **Scripts and Source**  
Rather than writing line-by-line, it is easier to create functions in a text editor and then copy and paste them into R. Many text editors will auto-generate brackets, parentheses, etc., making the programming much easier.  

Within R, you can create script files (which have the .R extension) than can be called using the `source()` function. Rather than copy and pasting all of the text in a script, you can use the `source()` function to run an entire script.  

For example, execute the following script to plot a Christmas tree:
```{r,echo=TRUE,eval=FALSE}
source( "https://raw.githubusercontent.com/jacobtnyoung/RWorkshop/main/tree_source.R" )
```

Now, let's take a look at the file ([click here](https://raw.githubusercontent.com/jacobtnyoung/RWorkshop/master/tree_source.R)) and see what the script is doing.  
<br>

## **Loops**  
Often we need to repeat an action several times and would like a procedure that does this for us. In R, the `for()` function accomplishes this task and takes the following form:  
```{r,echo=TRUE,eval=FALSE}
for( i in 1:n ){  
  #do some stuff here  
  }  
```

In this syntax, `for` indicates that we are going to loop from a start index to an end index, `i` is the index we are looping over, `1` is our start index, `n` is the end index, `{` opens the loop and `}` closes the loop. With loops, actions that take place in the loop involve objects, and these objects must be created before we initiate the loop.  

Let’s create a loop that takes each value of an object and squares it. For example:  
```{r,echo=TRUE,eval=TRUE}
  x <- seq( 1,5,1 )	        #create an object called x that is 1,2,3,4,5.
  x.squared <- NULL       #create an object that is empty which we will use in the loop.
    for( i in 1:5 ){	       #for each element for 1 through 5.
      x.squared[i] <- x[i]^2 #each element of x.squared will be x[i] squared.
	}
x.squared # print the object to see the value.
```  

R also has some very helpful loops already programmed in the `apply()` function family. The `apply()` function allows us apply to a function to a two-dimensional object. The `apply()` function has three arguments: the object, whether to apply across rows or columns, the functions to use. Let's check out the help:  
```{r,echo=TRUE,eval=TRUE}
?apply
```

For example, say we have a data set with 5 individuals and 3 variables and we want the mean for each of the variables:

Person | Var1 | Var2 | Var3
------ | ---- | ---- | ---- 
John   | 1    | 2    | 3
Essa   | 4    | 5    | 5
Stan   | 4    | 3    | 2
Cora   | 1    | 1    | 2
Dezy   | 3    | 4    | 5
------ |      |      | 
Mean   | ?    | ?    | ?    

If we read these data in as a matrix, then the mean for each variable is just the column means:  
```{r,echo=TRUE,eval=TRUE}
#Create the data matrix using the matrix function.
dat <- matrix(
  c( 1,2,3,4,5,5,4,3,2,1,1,2,3,4,5 ), # a vector of the values.
  nrow=5, # the number of rows in the matrix.
  ncol=3, # the number of columns in the matrix.
  byrow=TRUE #instructions to read by row (vs. by column).
  )
```

Note, we could write:  
```{r,echo=TRUE,eval=FALSE}
mean( dat[,1] )  
mean( dat[,2] )  
mean( dat[,3] )
```

Or, we could just use the `apply()` function:  
```{r,echo=TRUE,eval=TRUE}
apply( dat,2,mean )
```
<br>

## **If/Else Statements**  
R has a useful function called `which()` that returns T/F statements about objects. The function returns the position of the values meeting some condition. For example:  
```{r,echo=TRUE,eval=TRUE}
x <- c( 1,2,3,4,5,5,4,3,2,1,1,2,3,4,5 )
which( x > 3 ) #which elements of x are greater than 3?
```
Shows that elements in the 4th, 5th, 6th, 7th, 14th, and 15th positions are > 3.

Using the same idea, the `if()` function can be used if we want to perform a certain task depending on some condition. The basic structure is:  
`if(`statement that gives TRUE or FALSE`) {`  
    action to take  
`}`  

For example, say we want to recode some value if it meets some condition. Specifically, lets recode all values in `x` to be 0 of they are greater than 3. To do this, we need to put our if statement in a `for()` loop.  
```{r,echo=TRUE,eval=TRUE}
x <- c( 1,2,3,4,5,5,4,3,2,1,1,2,3,4,5 )

for( i in 1:15 ){
    if( x[i] > 3 ){ #define the condition.
      x[i] <- 0 #assign the value if condition is met.
    }
  }
x
```  

There is also the `else` command that lets us perform actions if some condition is not met. For example, recode all values to be 1 if they are not greater than 3 (i.e. going to be changed to 0).  
```{r,echo=TRUE,eval=TRUE}
x <- c( 1,2,3,4,5,5,4,3,2,1,1,2,3,4,5 )
for( i in 1:15 ){
  if( x[i] > 3 ){
    x[i] <- 0
    }
  else x[i] <- 1
  }
x
```  

Finally, there is also the `else if()` function that lets you combine several possible actions. For example:  
```{r,echo=TRUE,eval=TRUE}
x <- c( 1,2,3,4,5,5,4,3,2,1,1,2,3,4,5 )

for( i in 1:15 ){
  if( x[i] > 3 ){ #recode 4 and 5 to 0.
    x[i] <- 0
    }
  else if( x[i] == 3 ){ #recode 3 to 2.
    x[i] <- 2
    }
  else x[i] <- 1 # recode 1 and 2 to 1.
  }
x
```
<br>

## **Questions?**  