---
title: "Data Visualization: R Graphics"
date: "R Workshop"
output: 
  html_document:
    df_print: paged
    theme: cosmo
    highlight: haddock
    toc: yes
    toc_float: yes
    code_fold: show
    self_contained: true
---

<br>

# **Introduction**

One of the best features of R is its graphics capability. As far as flexibility and breadth, R graphics is unsurpassed (in my opinion). There are MANY functions in R that involve visualizing data. There are many examples online, but a nice site that has tons of graphics (and syntax!) is: [http://www.r-graph-gallery.com/](http://www.r-graph-gallery.com).

<br>

## **Why Visualization matters...**

Pictures are cool, no doubt. But, visualization of information is an ESSENTIAL part of your toolkit as a researcher, analyst, and/or rad R user. Visualizing information helps you tell the story and make the point to those who are "visual" learners. 

For example, in a paper entitled ["Solitary Confinement and the Well-Being of People in Prison"](https://www.sciencedirect.com/science/article/pii/S0277953623005816), Wright et. al. examined the effects of solitary confinement on mental well-being at three time points over a 12-month period (called "baseline", "6-Month", and "12-Month"). A key feature of the study was tracking movement of individuals through custody levels.

<br>

Take a look at the table below which shows a cross-tabulation for custody levels for individuals. That is, the cells represent the count of individuals who were at a particular custody level at baseline (the row) and their custody level at 6 months (the column):

*Custody Levels*    | Minimum (BL) | Medium | Close | Maximum | Attrite |
|:-----------------:|:------------:|:------:|:-----:|:-------:|:-------:|
|   **Medium (6M)** |       6      |   81   |   3   |    3    |   17    |
|   **Close**       |       0      |    9   |  60   |   13    |   12    |
|   **Maximum**     |       0      |    0   |  14   |   99    |    9    |

*What does this table tell you?* In other words, what can you infer from the table?

<br>

It is a bit hard to follow. Now, what if we added the two other tables: 6-month to 12-month and baseline to 12-month. That would be a lot of information to digest and tough to get a sense of what was occurring.

A visualization would help. How could we better represent this information?

Take a look at [this visualization](../images/custody-flow-plot.html).

```{r sanky, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}

# ----
library( networkD3 )
library( dplyr )
library( here )

# ----
RH.dat <- read.csv( here( "custody-flow.csv" ), as.is = TRUE, header = TRUE )

# ----
# A connection data frame is a list of flows with intensity for each flow
links <- data.frame(
  source=c( 
    # Baseline senders.
    "Max Baseline"   , "Max Baseline"   , "Max Baseline"   , "Max Baseline"   , "Max Baseline"   ,
    "Close Baseline" , "Close Baseline" , "Close Baseline" , "Close Baseline" , "Close Baseline" ,
    "Medium Baseline", "Medium Baseline", "Medium Baseline", "Medium Baseline", "Medium Baseline",
    # 6 Month senders.
    "Max 6 Month"   , "Max 6 Month"   , "Max 6 Month"   , "Max 6 Month"   , "Max 6 Month"   ,
    "Close 6 Month" , "Close 6 Month" , "Close 6 Month" , "Close 6 Month" , "Close 6 Month" ,
    "Medium 6 Month", "Medium 6 Month", "Medium 6 Month", "Medium 6 Month", "Medium 6 Month",
    "Min 6 Month"   , "Min 6 Month"   , "Min 6 Month"   , "Min 6 Month"   , "Min 6 Month"
            ), 
  target=c( 
    # 6 Month receivers, from baseline senders.
    "Max 6 Month", "Close 6 Month", "Medium 6 Month", "Min 6 Month", "Not In 6 Month",
    "Max 6 Month", "Close 6 Month", "Medium 6 Month", "Min 6 Month", "Not In 6 Month",
    "Max 6 Month", "Close 6 Month", "Medium 6 Month", "Min 6 Month", "Not In 6 Month",
    # 12 Month receivers, from 6 Month senders.
    "Max 12 Month", "Close 12 Month", "Medium 12 Month", "Min 12 Month", "Not In 12 Month",
    "Max 12 Month", "Close 12 Month", "Medium 12 Month", "Min 12 Month", "Not In 12 Month",
    "Max 12 Month", "Close 12 Month", "Medium 12 Month", "Min 12 Month", "Not In 12 Month",
    "Max 12 Month", "Close 12 Month", "Medium 12 Month", "Min 12 Month", "Not In 12 Month"
            ), 
  value=c( 
    #
    sum( RH.dat$Max_6[ RH.dat$Max == 1 ] ), # max to max
    sum( RH.dat$Close_6[ RH.dat$Max == 1 ] ), # max to close
    sum( RH.dat$Med_6[ RH.dat$Max == 1 ] ), # max to med
    sum( RH.dat$Min_6[ RH.dat$Max == 1 ] ), # max to min
    sum( RH.dat$Max[ RH.dat$in_6mfollowup != 1 ] ), # max to out of sample.
    #
    sum( RH.dat$Max_6[ RH.dat$Close == 1 ] ), # max to close
    sum( RH.dat$Close_6[ RH.dat$Close == 1 ] ), # close to close
    sum( RH.dat$Med_6[ RH.dat$Close == 1 ] ), # close to med
    sum( RH.dat$Min_6[ RH.dat$Close == 1 ] ), # close to min
    sum( RH.dat$Close[ RH.dat$in_6mfollowup != 1 ] ), # close to out of sample.
    #
    sum( RH.dat$Max_6[ RH.dat$Med == 1 ] ), # med to close
    sum( RH.dat$Close_6[ RH.dat$Med == 1 ] ), # med to close
    sum( RH.dat$Med_6[ RH.dat$Med == 1 ] ), # med to med
    sum( RH.dat$Min_6[ RH.dat$Med == 1 ] ), # med to min
    sum( RH.dat$Med[ RH.dat$in_6mfollowup != 1 ] ), # med to out of sample.
   ##
    #
    sum( RH.dat$Max_12[ RH.dat$Max_6 == 1 ] ), # max to max
    sum( RH.dat$Close_12[ RH.dat$Max_6 == 1 ] ), # max to close
    sum( RH.dat$Med_12[ RH.dat$Max_6 == 1 ] ), # max to med
    sum( RH.dat$Min_12[ RH.dat$Max_6 == 1 ] ), # max to min
    sum( RH.dat$Max_6[ RH.dat$in_12mfollowup != 1 ] ), # max to out of sample.
    #
    sum( RH.dat$Max_12[ RH.dat$Close_6 == 1 ] ), # max to close
    sum( RH.dat$Close_12[ RH.dat$Close_6 == 1 ] ), # close to close
    sum( RH.dat$Med_12[ RH.dat$Close_6 == 1 ] ), # close to med
    sum( RH.dat$Min_12[ RH.dat$Close_6 == 1 ] ), # close to min
    sum( RH.dat$Close_6[ RH.dat$in_12mfollowup != 1 ] ), # close to out of sample.
    #
    sum( RH.dat$Max_12[ RH.dat$Med_6 == 1 ] ), # med to close
    sum( RH.dat$Close_12[ RH.dat$Med_6 == 1 ] ), # med to close
    sum( RH.dat$Med_12[ RH.dat$Med_6 == 1 ] ), # med to med
    sum( RH.dat$Min_12[ RH.dat$Med_6 == 1 ] ), # med to min
    sum( RH.dat$Med_12[ RH.dat$in_12mfollowup != 1 ] ), # med to out of sample.
    #
    sum( RH.dat$Max_12[ RH.dat$Min_6 == 1 ] ), # min to close
    sum( RH.dat$Close_12[ RH.dat$Min_6 == 1 ] ), # min to close
    sum( RH.dat$Med_12[ RH.dat$Min_6 == 1 ] ), # min to med
    sum( RH.dat$Min_12[ RH.dat$Min_6 == 1 ] ), # min to min
    sum( RH.dat$Min_6[ RH.dat$in_12mfollowup != 1 ] ) # min to out of sample.
  )
)

# ----
# From these flows we need to create a node data frame: it lists every entities involved in the flow
nodes <- data.frame(
  name=c( as.character( links$source ), 
         as.character( links$target ) ) %>% unique()
)

# ----
# With networkD3, connection must be provided using id, not using real name like in the links dataframe. So we need to reformat it.
links$IDsource <- match( links$source, nodes$name )-1 
links$IDtarget <- match( links$target, nodes$name )-1


# ----
# Add a group that shows movement through custody levels.
links$custody.moves <- as.factor( c( 
  "same",  "down",  "down",  "down",  "out",
  "up",  "same",  "down",  "down",  "out",
  "up",  "up",  "same",  "down",  "out",
  "same",  "down",  "down",  "down",  "out",
  "up",  "same",  "down",  "down",  "out",
  "up",  "up",  "same",  "down",  "out",
  "up",  "up",  "up",  "same",  "out"
 )
)

# ----
# Add a group that shows how people in specific custody levels move:
links$custody.init <- as.factor( c( 
  "Max","Max","Max","Max","Max",
  "Close","Close","Close","Close","Close",
  "Medium","Medium","Medium","Medium","Medium",
  "Max","Max","Max","Max","Max",
  "Close","Close","Close","Close","Close",
  "Medium","Medium","Medium","Medium","Medium",
  "Min","Min","Min","Min","Min"
 )
)


# Remove the flows with zero transfers.
links <- links[links$value != 0,]

# Add a 'group' column to each node. Here I decide to put all of them in the same group to make them grey
nodes$custody.moves <- as.factor(c("custody.moves"))
nodes$custody.init <- as.factor(c("custody.init"))

# Give a color for each group:
color.moves <- 'd3.scaleOrdinal() .domain(["same", "up", "down", "out", "custody.moves"]) .range(["#b6ceed", "#da6763", "#5ddaab", "#e0c380", "grey"])'
color.init <- 'd3.scaleOrdinal() .domain(["Max", "Close", "Medium", "Min", "custody.init"]) .range(["#942823", "#cb8630", "#24759a", "#a627a2", "grey"])'

# ----
# Make the network.
p <- sankeyNetwork( Links = links, Nodes = nodes,
                    Source = "IDsource", Target = "IDtarget",
                    Value = "value", NodeID = "name", 
                    sinksRight=FALSE, fontSize = 12,
                    colourScale=color.moves, LinkGroup="custody.moves", NodeGroup="custody.moves"
                    )
p

#library( htmlwidgets )
#saveWidget(p, paste0( path, "/custody-flow-plot.html" ) )

```

### The Point

Visualization is a great medium for conveying information. There are a LOT of tools in R for visualizing information. So, to get a sense of how it all works, let's start with a basic function `plot()`.

<br>

## **The `plot()` Function**

A very useful graphics function is the `plot()` function. This function plots a two-dimensional pane with two arguments giving the x and y coordinates.

Let's create a simple plot:

```{r,echo=FALSE,eval=TRUE}

set.seed( 12345 )

```

```{r,echo=TRUE,eval=TRUE}

x <- rnorm( 100, 0, 1 ) #create a vector with 100 elements drawn from a normal distribution.
y <- seq( 1, 10, length.out=length( x ) ) # create a vector 1:10 with same length.
plot( x, y ) #plot it.

```

This is a pretty simple plot. As mentioned, one of the nice features of R is the flexibility to create your plot. Take a look at the different arguments that we pass to `plot()` to modify it by examining the help: `?plot`.

```{r, results='asis', echo=FALSE}

cat( "\\newpage" )

```

<br>

Overall, there are *many* different parameters we can modify in `plot()`. Let's check out a few:

<br>


### Type of plot

We can change the "type" of plot:

```{r,echo=TRUE,eval=FALSE}

plot( x, y, type="l" ) #plot a line.
plot( x, y, type="p" ) #plot points.
plot( x, y, type="b" ) #plot both!

```

Often, when plotting multiple objects, we want to first set up the plot regions before adding anything. This is a plot of type "none": `plot( x,y, type="n" )`.

```{r,echo=TRUE,eval=TRUE}

plot( x, y,
   type="n",
   main="our sample plot", # plot a title.
   xlab="this is the x axis", # change the x label.
   ylab="this is the y axis" # change the y label.
  )

```

<br>

### Characters

We can also change the "characters" of the plot:

```{r,echo=TRUE,eval=FALSE}

plot( x, y, pch=1 ) #plot a point.
plot( x, y, pch=2 ) #plot a triangle.
plot( x, y, pch=3 ) #plot a +.
plot( x, y, pch=4 ) #plot an x.

```


The argument **pch** determines the shape of the plot points. The numeric values 0 to 25 represent different default shapes. We can also use any number, letter, or symbol as a plotting shape.

```{r, echo=FALSE}

pchShow <-
  function(extras = c("*",".", "o","O","0","+","-","|","%","#"),
           cex = 3, ## good for both .Device=="postscript" and "x11"
           col = "darkgray", bg = "lightblue", coltext = "black", cextext = 1.2,
           main = paste("plot symbols :  points (...  pch = *, cex =",
                        cex,")"))
  {
    nex <- length(extras)
    np  <- 26 + nex
    ipch <- 0:(np-1)
    k <- floor(sqrt(np))
    dd <- c(-1,1)/2
    rx <- dd + range(ix <- ipch %/% k)
    ry <- dd + range(iy <- 3 + (k-1)- ipch %% k)
    pch <- as.list(ipch) # list with integers & strings
    if(nex > 0) pch[26+ 1:nex] <- as.list(extras)
    plot(rx, ry, type = "n", axes  =  FALSE, xlab = "", ylab = "", main = main)
    abline(v = ix, h = iy, col = "lightgray", lty = "dotted")
    for(i in 1:np) {
      pc <- pch[[i]]
      ## 'col' symbols with a 'bg'-colored interior (where available) :
      points(ix[i], iy[i], pch = pc, col = col, bg = bg, cex = cex)
      if(cextext > 0)
          text(ix[i] - 0.3, iy[i], pc, col = coltext, cex = cextext)
    }
  }

pchShow()

```

Note that shapes 0 to 14 are hollow, 15 to 20 are solid, and 21 to 25 can also plot a background color specified by the **bg=** argument.

<br>

### Points, lines, text, etc.

Additionally, the `points()`, `lines()`, `segments()`, and `text()` functions are useful for adding information to plots.

```{r, results='asis', echo=FALSE}

cat("\\newpage")

```

Here is an example I use in my data analysis course to illustrate the properties of the standard normal distribution.

First, let's set up our values:

```{r,echo=TRUE,eval=TRUE}

y    <- seq( -15, 30, length=1000 ) # sequence from -15 to 30. 
hx.1 <- dnorm( y, 0, 1 ) # densities for the plots.
hx.2 <- dnorm( y, 0, 2 )
hx.3 <- dnorm( y, 0, 3 )

```

Next, let's set up the plot, but we don't want to add anything yet (so we use `type="n"`):

```{r,echo=TRUE,eval=TRUE}

plot( y, hx.1,
  xlab="", ylab="", # blank out the labels for x and y.
  type="n", #do not plot anything.
  main="Normal Distributions" # a title.
)

```

Now, illustrate the shape of the distributions using the `lines()` function (you can copy and paste one at a time to see them get added):

```{r,echo=TRUE,eval=TRUE}

plot( y, hx.1,
  xlab="", ylab="", # blank out the labels for x and y.
  type="n", #do not plot anything.
  main="Normal Distributions" # a title.
)

lines( y, hx.1, col="blue", type="l", lwd=2 )
lines( y, hx.2, col="red", type="l", lwd=2 )
lines( y, hx.3, col="darkgreen", type="l", lwd=2 )

```
  
Now, add a line to show the central tendency by using the `segments()` function:

```{r,echo=TRUE,eval=TRUE}

plot( y, hx.1,
  xlab="", ylab="", # blank out the labels for x and y.
  type="n", #do not plot anything.
  main="Normal Distributions" # a title.
)

lines( y, hx.1, col="blue", type="l", lwd=2 )
lines( y, hx.2, col="red", type="l", lwd=2 )
lines( y, hx.3, col="darkgreen", type="l", lwd=2 )

segments( 0, 0, 0, 0.5, col="black", lwd=2 )

```

Finally, add some text to show the values (note that we will use the `text()` function):

```{r,echo=TRUE,eval=TRUE}
plot( y, hx.1,
  xlab="", ylab="", # blank out the labels for x and y.
  type="n", #do not plot anything.
  main="Normal Distributions" # a title.
)

lines( y, hx.1, col="blue", type="l", lwd=2 )
lines( y, hx.2, col="red", type="l", lwd=2 )
lines( y, hx.3, col="darkgreen", type="l", lwd=2 )

segments( 0, 0, 0, 0.5, col="black", lwd=2 )

text( 11, 0.35, "Mean = 0, SD = 1", col="blue", cex=1.5 )
text( 12, 0.15, "Mean = 0, SD = 2", col="red", cex=1.5 )
text( 13, 0.06, "Mean = 0, SD = 3", col="darkgreen", cex=1.5 )
```

```{r, results='asis', echo=FALSE}

cat("\\newpage")

```


## **The Layering Approach**  

As we have seen, we can start with a basic plot and add information. Creating graphics in this way is referred to as *layering* because we are stacking additional layers of elements on top of each other.  

Consider the following plot:  

```{r,echo=FALSE,eval=TRUE}
  x<-c(1960,1965,1970,1975,1980,1985,1990,1995)
  y<-c(0.18,0.20,0.24,0.30,0.47,0.61,0.70,0.90)

# Set up the plot.    
  plot(x, y,
	  xlim=c(min(x)-5,max(x)+5),
		ylim=c(min(y)-0.5, max(y)+0.5),
		main="Plot of Average Family Deaths by Year",
		xlab="Year",
		ylab="Average Family Deaths",
		pch = 21, col = "red", bg = "lightblue"
  )

# Add the least-squares regression line to the plot.  
  abline(lm(y~x),lty=2) #the function lm() is the linear model of y regressed on x. 
   
# Add some more features.
  points(mean(x),mean(y),col="black",pch=3,cex=3) # add some points to the plot.
	abline(h=mean(y),lty=3) #plot the mean of y horizontally.
	abline(v=mean(x),lty=3) #plot the mean of x vertically.

  segments(1985,mean(y),1985,0.61,lwd=3,col="red")
	text(1987.2,0.53,"y-ybar")
	segments(mean(x),0.61,1985,0.61,lwd=3,col="red")
	text(1981,0.65,"x-xbar")
```

As you can see, there are a number of elements that have been used to create this plot:  

  + a basic plot layout using `plot()`
  + a set of points using `points()`
  + several segment lines using `segments()`
  + a title and axis labels using the `main=`, `xlab=`, and `ylab=` **arguments** in the function `plot()`
  + and some words using `text()`.

Let's go through and build this plot, layer by layer.

<br>

First, what are these data? 

The data are yearly rates of family deaths recorded by a professor at Penn State. That is, the rate at which family deaths are reported to him prior to an exam from 1960-1995. 

Here are what the data look like in a table:

| Year | Death Rate |
|:----:|:----------:|
| 1960 | 0.18       |
| 1965 | 0.20       |
| 1970 | 0.24       |
| 1975 | 0.30       |
| 1980 | 0.47       |
| 1985 | 0.61       |
| 1990 | 0.70       |
| 1995 | 0.90       |

<br>

Now, let's move the data into objects to work with in R:

```{r,echo=TRUE,eval=TRUE}
  x <- c( 1960, 1965, 1970, 1975, 1980, 1985, 1990, 1995 )
  y <- c( 0.18, 0.20, 0.24, 0.30, 0.47, 0.61, 0.70, 0.90 )
```

Next, let's set up the plot and define the limits of the axes using `xlim=` and `ylim=`, define the title using `main=`, and set the axis labels using `xlab=` and `ylab=`:

```{r,echo=TRUE,eval=TRUE}
# Set up the plot.    
plot(x, y,
     xlim=c( min( x ) - 5, max( x ) + 5 ), # set x axis limits.
     ylim=c( min( y ) - 0.5, max( y ) + 0.5) , # same for y axis.
     main="Plot of Average Family Deaths by Year", # title.
     xlab="Year", # x axis label.
     ylab="Average Family Deaths", # same for y axis.
     type="n" # don't plot anything inside.
     )
```

Now, we can add the points to the plot using the `points()` function. We can customize the points using the `pch=`, `col=`, and `bg=` arguments.

```{r,echo=TRUE,eval=TRUE}
plot(x, y,
     xlim=c( min( x ) - 5, max( x ) + 5 ), # set x axis limits.
     ylim=c( min( y ) - 0.5, max( y ) + 0.5) , # same for y axis.
     main="Plot of Average Family Deaths by Year", # title.
     xlab="Year", # x axis label.
     ylab="Average Family Deaths", # same for y axis.
     type="n" # don't plot anything inside.
)

points( x, y, pch = 21, col = "red", bg = "lightblue" )
```

What additional information could we add to this plot that would aid in our understanding of the relationship between year and average family deaths? What about understanding how regression works?  

Well, we could add the least-squares regression line to the plot using the `abline()` function and the `lm()` function: 

```{r,echo=TRUE,eval=TRUE}
plot(x, y,
     xlim=c( min( x ) - 5, max( x ) + 5 ), # set x axis limits.
     ylim=c( min( y ) - 0.5, max( y ) + 0.5) , # same for y axis.
     main="Plot of Average Family Deaths by Year", # title.
     xlab="Year", # x axis label.
     ylab="Average Family Deaths", # same for y axis.
     type="n" # don't plot anything inside.
)

points( x, y, pch = 21, col = "red", bg = "lightblue" )

abline( lm( y ~ x ), lty=2 )
```

Additionally, we can illustrate how OLS estimation works. Recall that OLS finds the line that minimizes the sum of squared residuals. We can show that using the `points()`, `abline()`, `segments()`, and `text()` functions:

```{r,echo=TRUE,eval=TRUE}
plot(x, y,
     xlim=c( min( x ) - 5, max( x ) + 5 ), # set x axis limits.
     ylim=c( min( y ) - 0.5, max( y ) + 0.5) , # same for y axis.
     main="Plot of Average Family Deaths by Year", # title.
     xlab="Year", # x axis label.
     ylab="Average Family Deaths", # same for y axis.
     type="n" # don't plot anything inside.
)

points( x, y, pch = 21, col = "red", bg = "lightblue" )
abline( lm( y ~ x ), lty=2 )

# add some points to the plot.
points( mean( x ), mean( y ), col="black", pch=3, cex=3 )

# plot the mean of y horizontally.
abline( h=mean( y ), lty=3 )

#plot the mean of x vertically.
abline( v=mean( x ), lty=3 )

# add segments and text showing the deviations.
segments( 1985, mean( y ), 1985, 0.61, lwd=3, col="red" )
text( 1987.2, 0.53, "y-ybar" )
segments( mean( x ), 0.61, 1985, 0.61, lwd=3, col="red" )
text( 1981, 0.65, "x-xbar" )

```

<br>

## **This is just the beginning!!!**  

There is MUCH more you can do with just the `plot()` function. See `help( par )` for a list of all the arguments and options for plotting.

There are also many other options for plots in R. There are entire packages created for plotting. One in particular is the `ggplot2`. Check out this page showing [crime in Phoenix](https://jacobtnyoung.github.io/OpenCrimPhx/crime-trends.html).


<br>

## **Questions?**
  
____

<br>

<p align="center">
[Back to R Workshop page](https://jacobtnyoung.github.io/RWorkshop/)
</p>

<br>
  
##### ***Please*** report any needed corrections to the [Issues](https://github.com/jacobtnyoung/RWorkshop/issues) page. Thanks!
  
<br>
<br>
  
###### ***Last updated `r format(Sys.time(), '%d %B, %Y')`***