---
title: "Dynamic Visualization: R Shiny"
date: "R Workshop"
output: 
  html_document:
    df_print: paged
    theme: cosmo
    highlight: haddock
    toc: yes
    toc_float: yes
    code_fold: show
    self-countained: false
---

<br>

# **Introduction**  
Another great feature of RStudio is that you can create dynamic visualizations using R Shiny. Dynamic graphics allow a user to select parameters that change the visualization in some way. Graphics will update in real-time within a web browser.

<br>

#### Example

Recall earlier how we created the figure below using `plot()` function:

<br> 

```{r, echo=FALSE, eval=TRUE}

y    <- seq( -15, 30, length=1000 ) # sequence from -15 to 30. 
hx.1 <- dnorm( y, 0, 1 ) # densities for the plots.
hx.2 <- dnorm( y, 0, 2 )
hx.3 <- dnorm( y, 0, 3 )

plot( y, hx.1,
  xlab="", ylab="", # blank out the labels for x and y.
  type="n", #do not plot anything.
  main="Normal Distributions" # a title.
)

lines( y, hx.1, col="blue", type="l", lwd=2 )
lines( y, hx.2, col="red", type="l", lwd=2 )
lines( y, hx.3, col="darkgreen", type="l", lwd=2 )

segments( 0, 0, 0, 0.5, col="black", lwd=2 )

text( 11, 0.35, "Mean = 0, SD = 1", col="blue", cex=1.5 )
text( 12, 0.15, "Mean = 0, SD = 2", col="red", cex=1.5 )
text( 13, 0.06, "Mean = 0, SD = 3", col="darkgreen", cex=1.5 )

```

```{r, results='asis', echo=FALSE}

cat("\\newpage")

```

<br>

This figure is great and all, but what it would be a better learning exercise if the individual could manipulate the values.

Like this:





# **Render Functions**

TEXT

<br>

## **Questions?**
  
____

<br>
  
#### ***Please*** report any needed corrections to the [Issues](https://github.com/jacobtnyoung/RWorkshop/issues) page. Thanks!
  
<br>
<br>
  
###### ***Last updated `r format(Sys.time(), '%d %B, %Y')`***