---
title: "Working with Data"
date: "R Workshop"
output: 
  html_document:
    df_print: paged
    theme: cosmo
    highlight: haddock
    toc: yes
    toc_float: yes
    self_contained: true
---

<style>
body {
text-align: justify}
</style>

# **Introduction**

Data are read into R using several functions written for this purpose. In my experience .csv format is the most well behaved.

The `read.csv()` function is used to assign a data file to an object:\
`data <- read.csv("`the path and file.csv`",`other arguments here`)`

Let's take a look at the help: `?read.csv`.

There is a more general function, `read.table`, that can read different types of files. But, it requires more programming. In addition, there is a package called `foreign` that has functions for reading files from other programs (e.g. `read.spss` reads SPSS files; `read.dta` reads Stata files, etc.). However, in my experience, if you are sending something from Stata to R, for example, it is better to just have Stata write a .csv file using: the `outsheet` function in Stata (type `help outsheet` in Stata for information).\
<br>

## **Example Data File**

Let's work through an example. Start by downloading the "R_Workshop_data.csv" file from this [link](https://github.com/jacobtnyoung/RWorkshop/blob/main/r-workshop-data.csv) on the [RWorkshop website](https://jacobtnyoung.github.io/RWorkshop/).

The file contains 52 individuals and 4 variables. The variables are: respondents id ("id"), a binary variable indicating whether the respondent is male or female ("male" where "1" is male), a measure of the respondent's age ("age"), and a measure of risky behaviors engaged in by the respondent ("risky").

<br>

## **Loading Data Files**

Let's go ahead and import it into R:

First, set the directory where the file is by using the `setwd("/…")` function. If you're a windows user, replace every `"\"` in the file path with `"\\"`.

```{r,echo=TRUE,eval=FALSE}
#read in the data file.

dat <- read.csv(
    "r-workshop-data.csv", # the data file.
    header = TRUE,         # tell R to read the first row as variable names.
    as.is = TRUE,          # tell R to not make any conversions.
    na.strings = "."       # tell R that missing values are periods.
    )

dat #look at the object.

```

<br>

If you had a REALLY big data set, you could use the `head()` function or the `tail()` function to look at the first few (or last few) cases.

```{r,echo=TRUE,eval=FALSE}

head( dat )
tail( dat )

```

<br>

If you use the `View()` function, you can open the spreadsheet to see how R views the file and why we give it particular instructions.

```{r,echo=TRUE,eval=FALSE}

View( dat )

```

<br>

Note that we could *skip* the step of storing the .csv file locally and simply call file from the website url. For example:

```{r,echo=TRUE,eval=FALSE}

dat <- read.csv(
    "https://raw.githubusercontent.com/jacobtnyoung/RWorkshop/main/r-workshop-data.csv",
    header=TRUE, as.is=TRUE, na.strings="." #all the other arguments remain the same.
    )

dat #look at the object.

```

<br>

## **Working with Objects**

Now, we can work with the `dat` object. As we saw before with matrices, we can index parts of the object by referring to the dimensions:

```{r,echo=TRUE,eval=FALSE}

dim( dat ) # what are the dimensions of the object?
dat[,1] # first column or first variable.
dat[1,] # first row or first case.
dat[,c( 1,2 )] # first two columns.
dat[,1:2] # same thing as dat[,c(1,2)].
dat[1:25,] # just the first 25 cases.
dat[,-c(1)] # remove the first column.

```

<br>

We can also refer to the variables (i.e. columns) by their names enclosed in `""` or by using the `$` key.

```{r,echo=TRUE,eval=FALSE}

dat[,"id"] # returns the entire column named "id".
dat$id # returns the column called "id" within the object "data".

```

<br>

Referencing parts of an object in this way makes it easy to execute functions. For example, the function `summary()` provides the range, mean, median, and info about missing values:

```{r,echo=TRUE,eval=FALSE}

summary( dat ) #summarize all the variables.
summary( dat$age ) #just summarize age.
summary( dat[,"age"] ) #same thing, different syntax.
summary( dat[,c( "age","male" )] ) #just age and male.

```

<br>

Let's read into the data, but exclude the header to see how to assign names:

```{r,echo=TRUE,eval=FALSE}

#toggle the header argument by setting it to FALSE.
dat <- read.csv( "r_workshop_data.csv" , header = FALSE, as.is = TRUE)

dat     # look at the object, note the first row.
dat[1,] # R has created the variable names for us.

#we can see this from the colnames() function.
colnames( dat )

#change the names.
names <- c( "id","male","age","risky" ) # create an object of names.
colnames( dat ) <- names #tell R what the column names are.

```

<br>

Obviously, if we had the names we would read them in with `header=TRUE` in the `read.csv` line.

<br>

## **Merging Data Files**

We can merge two objects (or datasets) with the `merge()` function. For example, let's create two datasets where the cases do not match exactly (i.e. some cases are missing data on a merged variable) by using the `data.frame()` function:

```{r,echo=TRUE,eval=FALSE}
# Create the datasets (note the difference between them).
dat.a <- data.frame( id = c( 1,2,3,4,5 ), age = c( 10,12,14,15,11 ) )
dat.b <- data.frame( id = c( 1,2,3,5 )  , sex = c( "m","m","f","m" ) )

class( dat.a ) # object "data.a" is of class data.frame.

?merge # note the different merging options.

# We can now merge these two datasets to get a single object.
dat.c <- merge( dat.a,dat.b,by.x = "id" )
dat.d <- merge( dat.a,dat.b,by.x = "id", all.x=TRUE ) #note difference.
```

<br>

## **Exporting Data Files**

We can also write data files out for analysis in another program. For example, say we have created a variable in our data set that we could only create in R and we want to export the file to another program. For .csv files, we can just use the `write.csv` function. As with reading files, there is a more general function, `write.table`, that can write different types of files (see `?write.table`). For `write.csv` we just tell it what object we want to write out to a file and the file name:

First, set the directory where the file will go by using `setwd("/…")`. Then:

```{r,echo=TRUE,eval=FALSE}

write.csv( dat,"New_R_Workshop_data.csv" ) #file path can be added also.

```

<br>

## **Saving Data Files**

Recall that to save the workspace use the `save.image()` function. This function requires a file path, a file name, and the extension ".RData" which is the format for an R workspace file. Since our imported data is an object (i.e. `data`), it will be saved to the workspace:

First, set the directory where the file will go by using `setwd("/…")`. Then:

```{r,echo=TRUE,eval=FALSE}

save.image( "dat.RData" ) #file path can be added also.
rm( list=ls() ) # clear the workspace.
ls() # check the contents of the workspace.
load( "dat.RData" ) #load the workspace that was previously saved.

```

<br>

## **Questions?**
  
____

<br>

<p align="center">
[Back to R 2 Phoenix page](https://jacobtnyoung.github.io/R2PhX/)
</p>

<br>

#### ***Please*** report any needed corrections to the [Issues](https://github.com/jacobtnyoung/RWorkshop/issues) page. Thanks!
  
<br>
<br>
  
###### ***Last updated `r format(Sys.time(), '%d %B, %Y')`***